The classical mutual information in mean-field spin glass models

### 2015-06-12 sk_model_v15.tex
revised-6

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[t]
\includegraphics*[width=0.93\linewidth]{./draft_figs/Renyi_MC_v2}
\caption{ The classical disorder-averaged Renyi entropy per spin $[S_2(\omega)]/N$  
 in the S-K model on the $2$-sheets booklet with zero magnetic field: 
 $[S_2(\omega)]/N$ as a function of the inverse temperature $\beta$. The different panels 
 correspond to different booklet aspect ratios $\omega=3/4,1/2,1/4$ (see Fig.~\ref{cartoon}). 
 The triangles, squares and rhombi denote the Monte Carlo results for 
 booklets with $N=32,64,128$ spins per sheet. The plus symbols are the numerical 
 extrapolations to the thermodynamic limit. The dash-dotted lines are the  
 analytical results in the replica-symmetry (RS) approximation. 
 The circles are the analytical results obtained 
 using the one-step replica symmetry breaking (1-RSB) ansatz.
}
\label{Renyi-MC}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Figure~\ref{I2-MC} plots $[{\mathcal I}_2]/N$ versus $0\le\beta\le 2.5$ and $\omega=1/2,1/4,
1/8$ (panels from left to right in the Figure). Notice that by definition (cf. Eq.~\eqref{MI}) 
$[{\mathcal I}_n(\omega)]=[{\mathcal I}_n(1-\omega)]$. Circles, squares, and triangles  
are Monte Carlo data for $N=32,64,128$. In the high-temperature region $[{\mathcal I}_2]$ 
exhibits a vanishing behavior. Moreover, finite size effects are ``small''. Using Eq.~\eqref{logZ-ht} 
it is straightforward to derive the high-temperature behavior of $[{\mathcal I}_n]$ as  
%
\begin{equation}
[{\mathcal I}_n]=\frac{N\beta^2}{2}\omega(1-\omega)n+{\mathcal O}
(\beta^4).
\end{equation}
%
Interestingly, $[{\mathcal I}_2]/N$ increases upon lowering the temperature 
up to $\beta\approx 1$, where it exhibits a maximum. The height of this maximum 
decreases as a function of $\omega$. We numerically observe that its position 
is not simply related to the paramagnetic-glassy transition. 
%This is not surprising as $[{\mathcal I}_n]$ is constructed by combining the partition 
%functions on booklets with different aspect ratios $\omega$, $1-\omega$, $\omega=0$, and 
%$\omega=1$ (cf. Eq.~\eqref{MI}), and the critical temperature of the paramagnetic-glassy 
%transition is a non trivial function of $\omega$ (see Fig.~\ref{beta_c}). 
Furthermore, the 
data for $[{\mathcal I}_2]/N$ at different system sizes do not exhibit any crossing. This is 
in sharp contrast with local models~\cite{jaconis-2013}, where ${\mathcal I}_n/L$ exhibits a 
crossing at a second order phase transition. The dash-dotted line in Fig.~\ref{I2-MC} is the 
analytical result obtained using the replica symmetric (RS) approximation (see 
section~\ref{rs-section}). Formally, this is obtained using Eq~\eqref{MI} and Eq.~\eqref{rs-F}, 
and it is in perfect agreement with the MC data in the whole paramagnetic phase.

@@

To summarize, our approach for calculating the ratio of partition functions in 
Eq.~\eqref{ratio-trick} consists of four steps:
\begin{enumerate}
\item Do a full Monte Carlo sweep of the system on the booklet with $A=A_i$ 
to generate an importance sampled spin configuration. This can be done with any 
update scheme, such as standard Metropolis or cluster updates
\item For the spins living in $A_{i+1}-A_i$ calculate 
%
\begin{align}
W_a = \sum_i e^{-\beta E_i}.
\end{align}
%
Here (cf. Eq.~\eqref{sub-choice}) $i$ runs over all the $n2^4$ configurations of the 
four spins in $A_{i+1}-A_i$, and $E_i$ is the energy associated with configuration $i$. 
\item Calculate the quantity
\begin{align}
W_b = \sum_j e^{-\beta E_j}
\end{align}
where $j$ runs over all the $2^4$ configurations with the four spins living in $A_{i+1}-
A_i$ and different sheets are identified.
\item Calculate the transition probability $T_{i\rightarrow i+1}$ as
%
\begin{align}
T_{i \rightarrow i+1} = \Bigl< \frac{W_b}{W_a} \Bigr>.
\end{align}
\end{enumerate}
%
This method is much faster than simply counting configurations as it allows 
us to integrate over all possible configurations of the spins in $A_{i+1}-A_i$, 
treating the remaining ones as a bath that is updated by the regular Monte Carlo update. 
The computational cost of this procedure grows exponentially with the number of spins 
in $A_{i+1}-A_i$. 
%We also gain some speed by calculating the weights in each replica seperately, 
%and then recombining the results in a smart way, meaning that we only have to explicitly 
%loop over $2 \cdot 2^4 = 32$ configurations.
In our Monte Carlo simulations we calculate $Z(A_i,n,\beta)/Z(A_{i+1},n,\beta)$ 
for every disorder realization and $S_n(A)$ using Eq.~\eqref{renyi-app}. 
Finally, we average over the disorder to obtain $[S_n(A)]$.
%When calculating the mutual information
%\begin{align}
%I_{n}(A) = S_{n}(A) + S^*_{n}(N-A) - S_{n}(N)
%\end{align}
%we do this calculating for each specific realization of disorder, and then take the disorder average of this final quantity.
%The star in $S^*$ denotes the fact that even though $A$ denotes the size of the region that is connected, due to disorder one must be careful which specific spins are in the region.
%$S^*$ is defined in such a way that the connected spins in $S_{n}(A)$ and those in $S^*_{n}(N-A)$ contain every spin in the system exactly once.

%For each disorder and each $\omega_i$ we create a simulation, and measure
%\begin{align}
%\frac{Z(\omega_i,\alpha,\beta)}{Z(\omega_{i+1},\alpha,\beta)} = \\
%\frac{Z(\omega_i,\alpha,\beta)}{Z(\omega_{i-1},\alpha,\beta)}
%\end{align}

%Due to the infinite dimensional nature of the Sherrington-Kirkpatrick model, the calculation the R\'enyi mutual information is unable to exploit efficient methods that are available for finite dimensional systems [ref: JMS/Inglis].

%where $A$ changes only slightly, and build up the full quantity by taking the product of many such measurements.
%This method is usually referred to as the ratio trick, and in finite dimensional systems there are efficient ways to calculate this ratio\cite{stephan-2014}.
%In an infinite dimensional system the only efficient way to use this method is to increase $A$ more slowly than typical.
%For this study, $A$ is increased by only 4 sites at a time.

%First, if we want to sample from $A=0$ (two disconnected systems) to $A=N$ (two fully connected systems), let us use the ratio trick to write

@@

%\bibitem{bouchaud-2003}

%J.-P.~Bouchaud, F.~Krzakala, and O.~C.~Martin, Phys.\ Rev.\ B {\bf 68},

%224404 (2003).

%\bibitem{boettcher-2005}

%S.~Boettcher: Eur.\ Phys.\ J.\ {\bf B46}, 501 (2005).

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[t]
\includegraphics*[width=0.93\linewidth]{./draft_figs/Renyi_MC_v2}
\caption{ The classical disorder-averaged Renyi entropy per spin $[S_2(\omega)]/N$  
 in the S-K model on the $2$-sheets booklet with zero magnetic field: 
 $[S_2(\omega)]/N$ as a function of the inverse temperature $\beta$. The different panels 
 correspond to different booklet aspect ratios $\omega=3/4,1/2,1/4$ (see Fig.~\ref{cartoon}). 
 The triangles, squares and rhombi denote the Monte Carlo results for 
 booklets with $N=32,64,128$ spins per sheet. The plus symbols are the numerical 
 extrapolations to the thermodynamic limit. The inset in panel (a) shows 
 $[S_2]/N$ at fixed $\beta=2$, plotted versus $N^{-2/3}$.The dotted line is 
 a linear fit. In all the panels  the dash-dotted lines are the  
 analytical results in the replica-symmetry (RS) approximation. 
 The circles are the analytical results obtained 
 using the one-step replica symmetry breaking (1-RSB) ansatz.
}
\label{Renyi-MC}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Figure~\ref{I2-MC} plots $[{\mathcal I}_2]/N$ versus $0\le\beta\le 2.5$ and $\omega=1/2,1/4,
1/8$ (panels from left to right in the Figure). Notice that by definition (cf. Eq.~\eqref{MI}) 
$[{\mathcal I}_n(\omega)]=[{\mathcal I}_n(1-\omega)]$. Circles, squares, and triangles  
are Monte Carlo data for $N=32,64,128$. In the high-temperature region $[{\mathcal I}_2]$ 
exhibits a vanishing behavior. Moreover, finite size effects are ``small''. Using Eq.~\eqref{logZ-ht} 
it is straightforward to derive the high-temperature behavior of $[{\mathcal I}_n]$ as  
%
\begin{equation}
[{\mathcal I}_n]=\frac{N\beta^2}{2}\omega(1-\omega)n+{\mathcal O}
(\beta^4).
\end{equation}
%
Interestingly, $[{\mathcal I}_2]/N$ increases upon lowering the temperature 
up to $\beta\approx 1$, where it exhibits a maximum. The height of this maximum 
decreases as a function of $\omega$. We numerically observe that its position 
is not simply related to the paramagnetic-glassy transition. Furthermore, the 
data for $[{\mathcal I}_2]/N$ at different system sizes do not exhibit any crossing. This is 
in sharp contrast with local models~\cite{jaconis-2013}, where ${\mathcal I}_n/L$ exhibits a 
crossing at a second order phase transition. The dash-dotted line in Fig.~\ref{I2-MC} is the 
analytical result obtained using the replica symmetric (RS) approximation (see 
section~\ref{rs-section}). Formally, this is obtained using Eq~\eqref{MI} and Eq.~\eqref{rs-F}, 
and it is in perfect agreement with the MC data in the whole paramagnetic phase.

@@

To summarize, our approach for calculating the ratio of partition functions in 
Eq.~\eqref{ratio-trick} consists of four steps:
\begin{enumerate}
\item Do a full Monte Carlo sweep of the system on the booklet with $A=A_i$ 
to generate an importance sampled spin configuration. This can be done with any 
update scheme, such as standard Metropolis or cluster updates
\item For the spins living in $A_{i+1}-A_i$ calculate 
%
\begin{align}
W_a = \sum_i e^{-\beta E_i}.
\end{align}
%
Here (cf. Eq.~\eqref{sub-choice}) $i$ runs over all the $n2^4$ configurations of the 
four spins in $A_{i+1}-A_i$, and $E_i$ is the energy associated with configuration $i$. 
\item Calculate the quantity
\begin{align}
W_b = \sum_j e^{-\beta E_j}
\end{align}
where $j$ runs over all the $2^4$ configurations with the four spins living in $A_{i+1}-
A_i$ and different sheets are identified.
\item Calculate the transition probability $T_{i\rightarrow i+1}$ as
%
\begin{align}
T_{i \rightarrow i+1} = \Bigl< \frac{W_b}{W_a} \Bigr>.
\end{align}
\end{enumerate}
%
This method is much faster than simply counting configurations as it allows 
us to integrate over all possible configurations of the spins in $A_{i+1}-A_i$, 
treating the remaining ones as a bath that is updated by the regular Monte Carlo update. 
The computational cost of this procedure grows exponentially with the number of spins 
in $A_{i+1}-A_i$. 
In our Monte Carlo simulations we calculate $Z(A_i,n,\beta)/Z(A_{i+1},n,\beta)$ 
for every disorder realization and $S_n(A)$ using Eq.~\eqref{renyi-app}. 
Finally, we average over the disorder to obtain $[S_n(A)]$.

@@
